---
title: "scATAC-seq pre-processing"
author: "HDSU"
date: "19.05.2021"
vignette: >
  %\VignetteIndexEntry{scATAC-seq pre-processing}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    toc: yes
  #md_document:
  #  variant: markdown_github
editor_options: 
  chunk_output_type: console
---

  
```{r load_style, warning=FALSE, message=FALSE, results="hide"}
library(BiocStyle)
library(Seurat)
library(dplyr)
library(ArchR)
```

To compute the cell state-specific regulons (cssRegulons) the scATAC-seq must be processed beforehand to compute the location of TF motifs binding to open chromatin regions for each cell type. This set should be stored as a list of GRangesList:
  - Every element of the list is one of the cell types
  - For every cell type a GRangesList should be created with the location of binding motifs for all TFs.

GRAnet includes functions to extract the set of TF motifs locations binding to open chromatin regions from an **ArchRProject**. This Vignette shows the steps required to process the scATAC-seq data with ArchR.

# Example: Mouse atlas 8 tissues

## Create ArchR project

List of Arrow files and general settings:

```{r}
ArrowFiles  <- c("data/scatac/WholeBrainA_62216_barcoded.bam.arrow",
                 "data/scatac/Lung2_62216_barcoded.bam.arrow",
                 "data/scatac/Kidney_62016_barcoded.bam.arrow",
                 "data/scatac/BoneMarrow_62016_barcoded.bam.arrow",
                 "data/scatac/Liver_62016_barcoded.bam.arrow",
                 "data/scatac/Thymus_62016_barcoded.bam.arrow",
                 "data/scatac/Spleen_62016_barcoded.bam.arrow",
                 "data/scatac/SmallIntestine_62816_barcoded.bam.arrow")

ArchR::addArchRThreads(20)
addArchRGenome("mm9")
p_archrproj <- "data/scatac/archr/ArchRProject_/"

```


```{r}
#------------------------------------------------------------------------------#
#                                Create ArchR                                  #
#------------------------------------------------------------------------------#

# Creat ArchRProject
archrproj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = p_archrproj,
  copyArrows = TRUE 
)

archrproj
getAvailableMatrices(archrproj)



```

## Filter low quality cells

```{r}
#------------------------------------------------------------------------------#
#                      Filter low quality cells                                #
#------------------------------------------------------------------------------#

archrproj2 <- archrproj[log10(archrproj$nFrags) > 3.0, ]
hist(log10(archrproj$nFrags))
hist(log10(archrproj2$nFrags))


archrproj2 <- my_filterDoublets(archrproj2)
archrproj2

```


## Dimension reduction and clustering

```{r}
#------------------------------------------------------------------------------#
#                                  LSI                                         #
#------------------------------------------------------------------------------#

# Iterative Latent Semantic Indexing (LSI)
archrproj2 <- addIterativeLSI(
  ArchRProj = archrproj2,
  useMatrix = "TileMatrix",
  name = "IterativeLSI",
  iterations = 2,
  clusterParams = list( #See Seurat::FindClusters
    resolution = c(5),
    sampleCells = 10000,
    maxClusters = 20,
    n.start = 10
  ),
  varFeatures = 25000,
  dimsToUse = 1:30,
  UMAPParams = list(n_neighbors = 120, min_dist = 0.25)
)


# Clustering using Seurat’s FindClusters() function
archrproj2 <- addClusters(
  input = archrproj2,
  reducedDims = "IterativeLSI",
  method = "Seurat",
  name = "Clusters",
  resolution = 0.8
)


cM <- confusionMatrix(paste0(archrproj2$Clusters), paste0(archrproj2$Sample))

cM <- as.matrix(cM / Matrix::rowSums(cM))
p <- Heatmap(matrix = as.matrix(cM),
             col = viridis(100))
p

#------------------------------------------------------------------------------#
#                                  UMAP                                        #
#------------------------------------------------------------------------------#
# 6.1 Uniform Manifold Approximation and Projection (UMAP)
archrproj2 <- addUMAP(
  ArchRProj = archrproj2,
  reducedDims = "IterativeLSI",
  name = "UMAP",
  nNeighbors = 80,
  minDist = 0.25,
  metric = "cosine"
)

p1 <- plotEmbedding(ArchRProj = archrproj2, colorBy = "cellColData", name = "Sample", embedding = "UMAP")
p2 <- plotEmbedding(ArchRProj = archrproj2, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")

#------------------------------------------------------------------------------#
#                              Add tissue identity                             #
#------------------------------------------------------------------------------#
archrproj2$tissue <- paste0("Adult", gsub("_.*|Whole|2|A", "", archrproj2$Sample))
p3 <- plotEmbedding(archrproj2, colorBy = "cellColData", name = "tissue")

p1 + p2 + p3

```

Tissue distribution across clusters 

```{r}
h <- Heatmap(as.matrix(cM), col=magma(100), name="Confusion Matrix\nCluster vs Tissue")
h

```



##  Peak calling for each tissue

```{r}
#------------------------------------------------------------------------------#
#                   Pseudo-bulk Replicates in ArchR                            #
#------------------------------------------------------------------------------#
archrproj3 <- addGroupCoverages(ArchRProj = archrproj2, groupBy = "tissue")

#------------------------------------------------------------------------------#
#                         Calling Peaks with ArchR                             #
#------------------------------------------------------------------------------#
pathToMacs2 <- findMacs2()

archrproj3 <- addReproduciblePeakSet(
  ArchRProj = archrproj3,
  groupBy = "tissue",
  pathToMacs2 = pathToMacs2
)

archrproj3 <- addPeakMatrix(archrproj3)

```

## Motif binding location from cisbp database

```{r}
#------------------------------------------------------------------------------#
#                               Motif location                                 #
#------------------------------------------------------------------------------#
archrproj3 <- addMotifAnnotations(ArchRProj = archrproj3, motifSet = "cisbp", name = "Motif", force=TRUE)
getPositions(archrproj3)


```

## Save ArchR project

```{r}

saveArchRProject(ArchRProj = archrproj3,
                 outputDirectory = p_archrproj,
                 load = FALSE)


```






